---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
### Load the data (i.e. \color{red}{\verb|read.csv()|}read.csv())


The activity.zip file is located in the repository. The file activity.csv is loaded without unzipping.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)

# List files in .zip
unzip("activity.zip", list=TRUE)

# load .csv
data.raw <-read.table(unz("activity.zip", "activity.csv"), sep = ",", header = TRUE)

```

### Process/transform the data (if necessary) into a format suitable for your analysis

Exploring the data reveals the following
```{r}
str(data.raw)
```

Where the variables included in this dataset are:

**steps:** Number of steps taking in a 5-minute interval (missing values are coded as \color{red}{\verb|NA|}NA)  
**date:** The date on which the measurement was taken in YYYY-MM-DD format  
**interval:** Identifier for the 5-minute interval in which measurement was taken  
  
The date variable is turned into a date insted of factor.
```{r}
data.raw$date <- as.Date(data.raw$date)
str(data.raw)
```

## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

### Calculate the total number of steps taken per day
```{r results = "asis", message=FALSE}
data.naomit <- na.omit(data.raw)

total.steps <- data.naomit %>% group_by(date) %>% summarise(total_steps = sum(steps))
kable(total.steps, format = "html")

```


### If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day

```{r}
gg <- ggplot(total.steps)+
        geom_histogram(aes(total_steps), binwidth = 500)+
        ggtitle("Histogram of total steps taken")+
        xlab("Total steps")
gg
```


### Calculate and report the mean and median of the total number of steps taken per day
The mean is `r summary(total.steps$total_steps)[3]` and the median is `r summary(total.steps$total_steps)[2]`

```{r}
summary(total.steps$total_steps)
```



## What is the average daily activity pattern?

##Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}

intervals <- data.naomit %>% group_by(interval) %>% summarise(mean = mean(steps))

ggplot(intervals)+
        geom_line(aes(x=interval, y = mean))+
        ggtitle("5-minute intervals vs. mean amount of steps across days")

```



##Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
intervals %>% top_n(1, mean)
```


## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as \color{red}{\verb|NA|}NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

## Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs)
There are several ways to do this. I choose to create a column with the number of NA's for each row. I can then summarise the number of rows with one or more NA's.
```{r}
data.raw$row_nas <- rowSums(is.na(data.raw))
head(data.raw)

(n_nas <- data.raw %>% group_by(row_nas) %>% summarise(n_rows = n()))

```
There are `r n_nas[2,2]` rows with NA's.

## Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r}

# create column with mean for 5-minute interval
data.imp <- inner_join(data.raw, intervals)
head(data.imp)

# impute NA with the mean for 5-minute interval
data.imp$steps[is.na(data.imp$steps)] <- data.imp$mean[is.na(data.imp$steps)]
head(data.imp)

# remove temporary columns
data.imp <- data.imp[,1:3]

# tjek if there is any NA's
sum(is.na(data.imp))


```


## Create a new dataset that is equal to the original dataset but with the missing data filled in.
This is the data.imp data set.

```{r}
head(data.imp)
```


## Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r message = FALSE}
total.steps.imp <- data.imp %>% group_by(date) %>% summarise(total_steps = sum(steps))

gg.imp <- ggplot(total.steps.imp)+
        geom_histogram(aes(total_steps), binwidth = 500)+
        ggtitle("Steps taken with imputed data")+
        xlab("Total steps")


library(gridExtra)
grid.arrange(gg, gg.imp, nrow = 1)


summary(total.steps.imp$total_steps)
summary(total.steps$total_steps)

```
The result is that the distribution gets more narrow arround the mean, however as expectet the mean and median do not change. (actually the median changes by 1). This is expected becouse i used a imputation strategy where NA's was replaced by the mean from that 5-minute window.


## Are there differences in activity patterns between weekdays and weekends?

```{r}
Sys.setlocale("LC_TIME", "English")
table(weekdays(data.imp$date))

data.weekdays <- data.imp %>% filter(weekdays(date) != c("Saturday", "Sunday"))
data.weekends <- data.imp %>% filter(weekdays(date) == c("Saturday", "Sunday"))


total.steps.weekdays <- data.weekdays %>% group_by(date) %>% summarise(total_steps = sum(steps))
total.steps.weekends <- data.weekends %>% group_by(date) %>% summarise(total_steps = sum(steps))

hist.weekdays <- ggplot(total.steps.weekdays)+
        geom_histogram(aes(total_steps), binwidth = 500)+
        ggtitle("Weekdays total steps")+
        xlab("Total steps")
hist.weekends <-  ggplot(total.steps.weekends)+
        geom_histogram(aes(total_steps), binwidth = 500)+
        ggtitle("Weekends total steps")+
        xlab("Total steps")


intervals.weekdays <- data.weekdays %>% group_by(interval) %>% summarise(mean = mean(steps))
intervals.weekends <- data.weekends %>% group_by(interval) %>% summarise(mean = mean(steps))


time.weekdays <- ggplot(intervals.weekdays)+
        geom_line(aes(x=interval, y = mean))+
        ggtitle("Weekdays 5-min interval mean \nsteps across days")
time.weekends <- ggplot(intervals.weekends)+
        geom_line(aes(x = interval, y = mean))+
        ggtitle("Weekends 5-min interval mean \nsteps across days")

grid.arrange(hist.weekdays, hist.weekends, time.weekdays, time.weekends, nrow = 2)

summary(total.steps.weekdays$total_steps)
summary(total.steps.weekends$total_steps)

summary(intervals.weekdays$mean)
summary(intervals.weekends$mean)



```

The resulting plots and summries show that on avarage more steps are taken during weekdays, however in weeknds the steps are more evenly distributed across 5-min. intervals.